ODGI: understanding pangenome graphs

### 2021-10-31 manuscript/main.tex
added commads; moved view, matrix, paths, flatten, explode, squeezed, server, commands; multiple changes in results, extraction, editing, and metrics sections

\section{Implementation}
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:

@@

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

\subsection{Converting to standard file formats{\color{red} FIXME: We need a different title/subsection}
}
\label{sec:text}

\textit{odgi matrix} derives the pangenome matrix and writes the graph topology in sparse matrix formats.

\textit{odgi paths} allows the investigation of paths of a given variation graph. It can calculate overlap statistics of groupings of paths and extracts FASTA.

\textit{odgi flatten} projects the graph sequence and paths into FASTA and BED.

{\color{red} FIXME: We should put the following information in the metric section, of move some commands here}

\textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'.

\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.

\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently. {\color{red} FIXME: this command is listed in the list in 4 Results}

\textit{odgi degree/position/depth/untangle/tips} all can output the BED format.

\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.

\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
White spaces indicate where paths do not traverse the nodes.
The meaning of the colors depends on how \textit{odgi viz} is executed.
By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{b}).
Path names are displayed on the left of the paths.
The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig. \ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
Graph nodes order can affects downstream analyses on pangenome graphs. With \textit{odgi viz} we can color the paths by path position (Fig. \ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
Pangenome graphs represents both strand of DNA sequences.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig. \ref{fig:odgi_viz}\textbf{d}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig. \ref{fig:odgi_viz}\textbf{e}).

\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig. \ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}). A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.

\subsection{Extracting regions of interest}
\label{sec:extract}

% andrea (?)

% extract
Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
We can extract such regions using coordinates on the paths in the graph to guide us.

Figure showing extraction process---either schematic or ``real''.

odgi extract allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only with the extracted region.

%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.

%We extract key information about the graph using its paths as a pangenome reference system.

@@

% andrea (?)

% chop / unchop / squeeze / break / prune / explode

To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; \textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

Cycles in the graph complicate downstream analyses: \textit{odgi break} removes the cycles, reducing the complexity of the graph topology.
\textit{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most paths.

These are commonly-needed basic operations on the topology of the graph.

\subsection{Untangling and navigating the pangenome with path-based coordinates}
\label{sec:untangle}

@@

For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD! \citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
This approach updates node relative positions to best-match their distance in the paths running through the graph.
Based on this projection, we can trivially generate 1D and 2D visualizations.
We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.

In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig. \ref{fig:odgi_viz}\textbf{f-g}).
These allow for the direct translation of single coordinates and ranges, in an interactive way.
And, they let us extract a pairwise 1:many alignment between a given set of ``query'' sequences and a given set of ``target'' or reference sequences.
This lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.

@@

Figure showing the untangling of C4 (Fig. \ref{fig:odgi_viz}\textbf{h})
Figure showing tips over a single human chromosome from HPRC.

\subsection{Sorting the pangenome graph topology}
\label{sec:sort}

@@

% HTT exon1

% stats / multiQC / bin / degree / depth

@@

Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. It outputs pangenome statistics in TSV or YAML textual file formats.
MultiQC's \citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
ODGI also offers more advanced tools for the interrogation of the graphs. \textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. It provides tab delimited or JSON `bins'.
%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion
\textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
Collapsed sequences with highly identical repeats indicate intricate loci.
Complex motifs can be also detected with \textit{odgi degree}, which returns the node degree as defined by query criteria.
High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
In Fig. \ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.

@@

Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
%
Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder \citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

So far, ODGI has not been applied to graphs build from scWGS \citep{Zhuo2021} data. Each cell would be represented as a path in the graph, the path depth would be very high. As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.

>>>>

\section{Implementation}
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:

@@

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
\item \textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.~\ref{fig:odgi_viz}).
Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
White spaces indicate where paths do not traverse the nodes.
The meaning of the colors depends on how \textit{odgi viz} is executed.
By default, path colors has no particular meaning, but each path has its own color (Fig.~\ref{fig:odgi_viz}\textbf{b}).
Path names are displayed on the left of the paths.
The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
Graph nodes order can affects downstream analyses on pangenome graphs.
With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
Pangenome graphs represents both strand of DNA sequences.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).

\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.

\subsection{Extracting or joining regions of interest}
\label{sec:extract}

Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
such as a specific \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{a}) or a smaller region (Fig.~\ref{fig:odgi_viz}\textbf{b-g}), or even a single gene (Fig.~\ref{fig:metrics}),
thereby simplifying the downstream analyses and reducing the resources to work only with the extracted graphs.
%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process. For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.

In \textit{odgi extract}, we begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
Then, edges connecting all selected nodes are added in the subgraph under construction.
Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
Subpaths are search in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high depth.

% If needed, we can have a figure showing extraction process---either schematic or ``real''.

@@

Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping ~\citep{Eizenga_2020}.
However, graphs presenting complex topology can increase the computational overhead of many downstream analyses.
ODGI offers multiple commonly-needed basic operations on the topology of the graph and its nodes.

With \textit{odgi prune} users can remove complex parts as defined by query criteria, while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths.

To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}.
Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines~\citep{pggb}, but the current implementations return DAGs with 1-bp long nodes.
\textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

\subsection{Untangling and navigating the pangenome}
\label{sec:untangle}

@@

For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!~\citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
This approach updates node relative positions to best-match their distance in the paths running through the graph.
Based on this projection, we can trivially generate 1D and 2D visualizations.
We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.

In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).
These allow for the direct translation of single coordinates and ranges, in an interactive way.
And, they let us extract a pairwise 1:many alignment between a given set of ``query'' sequences and a given set of ``target'' or reference sequences.
This lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.

@@

Figure showing the untangling of C4 (Fig.~\ref{fig:odgi_viz}\textbf{h})
Figure showing tips over a single human chromosome from HPRC.

\FIXME{}
\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.

\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.

\textit{odgi position/untangle/tips} all can output the BED format.

\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
\FIXME{}

\subsection{Sorting the pangenome graph}
\label{sec:sort}

@@

@@

Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format.
Furthermore, \textit{odgi flatten} generate a linearization of the graph by emitting the pangenome sequence (the concatenation of all graph node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.

Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
It outputs pangenome statistics in tab-separated values (TSV) or YAML textual file formats.
MultiQC's~\citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
provides an interactive way to comparatively explore such statistics of an arbitrary number of graphs.

ODGI also offers more advanced tools for the interrogation of the graphs.
To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion

Genomes presenting sequences with highly identical repeats result in pangenome graphs with complex motifs that can be detected by \textit{odgi depth} and \textit{odgi degree}, which return the node depth and node degree, respectively, as defined by query criteria.
Both tools emit the output in BED format, allowing users to assess the complexity of the graph and detect intricate regions.
Indeed, high depth/degree nodes can be the mirror of genetic variation (Fig.~\ref{fig:metrics}), but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
%In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.

@@

Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
%
Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension~\citep{Wick_2015, Gonnella2018}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data. Each cell would be represented as a path in the graph, the path depth would be very high. As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.
